---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# IsletCalc

<!-- badges: start -->
[![R-CMD-check](https://github.com/sufyansuleman/IsletCalc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sufyansuleman/IsletCalc/actions/workflows/R-CMD-check.yaml)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
<!-- badges: end -->

IsletCalc computes surrogate indices of pancreatic **islet hormone release**
from fasting and 75 g oral glucose tolerance test (OGTT) measurements. It
covers both **beta-cell** (insulin) and **alpha-cell** (glucagon) function
in one lightweight, dependency-free package.

📖 **Full documentation, formulas and interpretation:**
<https://sufyansuleman.github.io/IsletCalc/>
([insulin release](https://sufyansuleman.github.io/IsletCalc/articles/insulin-release-indices.html) ·
[glucagon & the liver-alpha-cell axis](https://sufyansuleman.github.io/IsletCalc/articles/glucagon-liver-alpha-cell-axis.html)).

## Installation

```r
# development version:
# install.packages("remotes")
remotes::install_github("sufyansuleman/IsletCalc")
```

## Functions

| Function | Cell | Indices |
|---|---|---|
| `insulin_release()` | beta | HOMA-beta, CIR, Stumvoll, insulinogenic (Xinsdg30, Xinsg30), BIGTT-AIR, disposition (Di, Dibig) |
| `glucagon_release()` | alpha | fasting glucagon, glucagon AUC |
| `glucagon_resistance()` | alpha | glucagon suppression ratio, fasting glucagon:insulin ratio |

Beta-cell formulas follow Madsen et al. (2024, *Nature Metabolism*,
[doi:10.1038/s42255-024-01140-6](https://doi.org/10.1038/s42255-024-01140-6)).

## Example

```{r example}
library(IsletCalc)
data(example_data)

# beta-cell insulin release
res <- insulin_release(example_data, category = c("fasting", "ogtt"))
head(res[, c("Homa_beta", "Cir", "Bigtt_air", "Di")], 3)

# alpha-cell glucagon release and resistance
res <- glucagon_release(example_data)
res <- glucagon_resistance(res)
head(res[, c("Fasting_glucagon", "Glucagon_auc",
             "Glucagon_suppression_ratio")], 3)
```

A `Glucagon_suppression_ratio` at or above 1 indicates a failure to suppress
glucagon after the glucose load, a hallmark of hepatic glucagon resistance.

## Input data

| Column | Meaning | Unit |
|---|---|---|
| `G0`, `G30`, `G120` | Glucose at 0/30/120 min | mmol/L |
| `I0`, `I30`, `I120` | Insulin at 0/30/120 min | pmol/L |
| `sex` | Male indicator | 1 = male, else female |
| `bmi` | Body mass index | kg/m^2 |
| `Glucagon0`, `Glucagon30`, `Glucagon120` | Glucagon at 0/30/120 min | pmol/L |

Only the columns required for a requested index need be present; a missing
column skips that index with a warning rather than failing the whole call.

## Citation

```r
citation("IsletCalc")
```

## Related package

For insulin **sensitivity** indices (as opposed to release), see
[InsuSensCalc](https://github.com/sufyansuleman/InsuSensCalc).
