---
title: "Glucagon and the liver-alpha-cell axis"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(IsletCalc)
```

Where the beta cell secretes insulin, the pancreatic **alpha cell** secretes
**glucagon**. `glucagon_release()` and `glucagon_resistance()` summarize
alpha-cell output and its regulation from fasting and OGTT glucagon values.
This article explains the biology those indices are meant to capture.

## The liver-alpha-cell axis

Glucagon and hepatic amino-acid metabolism form a feedback loop known as the
**liver-alpha-cell axis**. Circulating amino acids stimulate glucagon
secretion and alpha-cell proliferation; glucagon, in turn, drives hepatic
amino-acid catabolism and ureagenesis. In health, an oral glucose load
**suppresses** glucagon, limiting hepatic glucose output.

**Glucagon resistance** describes a breakdown of this loop, seen with hepatic
steatosis and type 2 diabetes: the liver responds less to glucagon, amino
acids and glucagon both rise (hyperglucagonaemia), and glucose fails to
suppress glucagon normally. Non-suppression of glucagon after glucose is a
recognized early feature of dysglycaemia.

## Indices

### Glucagon release (`glucagon_release()`)

| Index | Definition | Meaning |
|---|---|---|
| `Fasting_glucagon` | `Glucagon0` | Basal alpha-cell tone; elevated in glucagon resistance. |
| `Glucagon_auc` | trapezoidal AUC over 0, 30, 120 min | Total glucagon exposure during the OGTT. |

The area under the curve uses the trapezoidal rule:
`0.5 x ((Glucagon30 + Glucagon0) x 30 + (Glucagon120 + Glucagon30) x 90)`.

### Glucagon resistance (`glucagon_resistance()`)

| Index | Definition | Meaning |
|---|---|---|
| `Glucagon_suppression_ratio` | `Glucagon120 / Glucagon0` | Below 1 = normal suppression after glucose; at or above 1 = failed suppression, a signature of glucagon resistance. |
| `Glucagon_insulin_ratio` | `Glucagon0 / I0` | Fasting alpha-to-beta balance; tracks whole-body insulin sensitivity and beta-cell function. |

For both ratios, glucagon and insulin should be supplied in a consistent
molar unit (for example pmol/L).

## Worked example

```{r}
data(example_data)
res <- glucagon_release(example_data)
res <- glucagon_resistance(res)
round(head(res[, c("Fasting_glucagon", "Glucagon_auc",
                   "Glucagon_suppression_ratio",
                   "Glucagon_insulin_ratio")], 5), 3)
```

Rows with `Glucagon_suppression_ratio` at or above 1 are the ones that failed
to suppress glucagon after the glucose load.

## References

- Chen X, Maldonado E, DeFronzo RA, Tripathy D (2021). Impaired Suppression
  of Glucagon in Obese Subjects Parallels Decline in Insulin Sensitivity and
  Beta-Cell Function. *J Clin Endocrinol Metab*, 106(5), 1398-1409.
  <https://doi.org/10.1210/clinem/dgab019>
- Wewer Albrechtsen NJ, et al. (2018). Evidence of a liver-alpha cell axis in
  humans: hepatic insulin resistance attenuates relationship between fasting
  plasma glucagon and glucagonotropic amino acids. *Diabetologia*, 61(3),
  671-680. <https://doi.org/10.1007/s00125-017-4535-5>
- Suppli MP, et al. (2020). Glucagon Resistance at the Level of Amino Acid
  Turnover in Obese Subjects With Hepatic Steatosis. *Diabetes*, 69(6),
  1090-1099. <https://doi.org/10.2337/db19-0715>
